---
output: pdf_document
---
Team4 - Duke Dinosaurs HW 1 Write Up

===================================================================================================

```{r}
source("graph.R")
```
VALID

Our is_valid function will test if the inputted graph object is a list of lists. We also check to see if the names of our vertices are unique, and that each vertex has a list of only edges and weights with edges being integers and weights being numerics. It will also check to make sure that no vertex connects to a nonexistent vertex and that each edge has a weight that is greater than 0. 

We write this function with if statements. Similar to a to-do list, we test conditions to be a valid graph. We start with the broadest conditions, the graph object has to be a list. We go on to test that each vertex is a list that contains edges and weights. The next step is to test if edges and weights are empty. If they aren't empty, the entry must not be NA. Then we test to make sure that edges are integers and weights are numerics. If our edges are greater than the length of our graph, or less than 1, there should be an error. This way, we can ensure that the number of vertices our vertex is connecting to is not more than the number of vertices in our graph and that it's not connecting to nothing. Our weights should not be less than or equal to 0. The number of vertices a vertex connects to should be the same number of weights that vertex holds, so each edge will have a weight. Lastly, we make sure that our vertex is only connecting to another vertex once. 
Let's test out an example graph. 

```{r}
graph1 <- list(A = list(edges   = c(2L, 5L, 6L),
                        weights = c(6, 1, 3)),
               B = list(edges   = c(1L, 3L, 6L),
                        weights = c(6, 4, 13)),
               C = list(edges   = c(2L, 6L),
                        weights = c(4, 7 )),
               D = list(edges   = c(5L, 6L),
                        weights = c(10, 12 )),
               E = list(edges   = c(1L,4L, 6L),
                        weights = c(1, 10, 5)),
               F = list(edges   = c(1L, 2L, 3L, 4L, 5L),
                        weights = c(3, 13, 7, 12, 5)))
is_valid(graph1)
```

is_valid(graph1) should return TRUE since as you can see from the graphic above, graph1 fits all the criterion of a valid graph. Since I know graph1 to be valid, it might be useful to run edge(graph1) just to have a better picture of what the graph looks like. 

For graphs such as the following, is_valid will return FALSE for various reasons. 

```{r}
g1 <- list(list(edges   = c(1L, 2L),
                     weights = c(1, 1)),
                list(edges   = c(1L, 2L),
                     weights = c(1, 1)))
is_valid(g1)
```
Our is valid function will return this graph input as FALSE because there aren't any unique names for the vertices. In fact, there aren't any names at all. 
 
```{r}
g2 = list(A = list(edges   = c(1L, 2L),
                         weights = c(1, 1)),
                B = list(edges   = c(1L, 3L),
                         weights = c(1, 1)))
is_valid(g2)
```
This will also return false, because if you look at vertex B$edges, you can see that it goes to 1L and 3L. 1L would be the first vertex in the list, so A but 3L is linking to the third vertex when it doesn't exist in g2. 

Undirected 

This function will test to see if a graph is undirected. A graph is undirected if all directed edges within the graph have a directed edge coming back with the same weight. For example, should vertex A connect to B with weight w, B should also connect to A with weight w. 

We first test to see if the inputted graph is valid. 

Afterwards, we utilize a helper function, edge, that will turn a graph into a dataframe, such that each row will represent a directed edge with the first column noting the starting vertex, the second column noting the ending vertex, and the third column noting the weight of that edge. 
```{r}
edge(graph1)
is_undirected(graph1)
```
Since we know that graph1 is valid, we can go ahead and show an example of what the edge function will do. Inputting graph1 into our is undirected function will also result in TRUE. Looking at the the dataframe we created using our edge function, you should be able to see that each row should have a complimentary row. We test for undirected-ness in our function using the following method. 

We only consider the case where a graph has more than 0 edges. Utilizing a for loop to run through all the rows in the graph's dataframe, we then use if statements to knock out graphs with attriubtes of an undirected graph. We first test to see if there's a row such that the ith row that we are testing will have the starting vertex as the ending vertex and vice-versa. If one does exist, we then test the weight of that row with our ith row. If the weight is the same, then we have a complimentary directed edge for our ith row.  

The following examples will be FALSE for a few different reasons. 
```{r}
g3 = list(A = list(edges   = c(1L,2L),
                     weights = c(1,1)),
            B = list(edges   = c(2L),
                     weights = c(1)))
is_undirected(g3)
```
This will return FALSE because as you can see A will go to itself with weight 1. B will go to itself with weight 1. However, A will also go to B with weight 1, but B doesn't go to A with weight 1. So the edge from A to B with weight 1 isn't reciprocated. 

```{r}
g4 = list(A = list(edges   = c(1L,2L),
                     weights = c(1,1)),
            B = list(edges   = c(1L,2L),
                     weights = c(2,1)))
is_undirected(g4)
```
This will also return FALSE. The issue here is that A goes to B with weight 1, but B goes to A with weight 2. 


Isomorphic

The purpose of this function is that given two graph objects, the function will give us TRUE if isomorphic, and FALSE if not. If two graphs are isomorphic, the vertices, edges, and weights are identical. Say we have vertex A in the first graph, and it connects to vertex B with weight w. We should also have a vertex A in the second graph that connects to vertex B with weight w. I call them vertex A and vertex B, because isomorphic graphs will be comparing the names of the vertices, not their position within the graph. This should hold true for the entire graph. 

First thing to note. We have created a helper function called edge. Edge takes a graph object and creates a data frame with three columns, start, end, and weight. The start and end columns will contain the name of the vertices. Weight will be a column of numbers. Each row will represent an edge and list the starting vertex, ending vertex and the weight. 

Aside from that, we utilized if statements and for loops to achieve our purpose. Subsetting and logical operators were also very helpful in eliminating graphs step by step. We worked from the broadest aspects we didn't want and then slowly became more detailed. 

The first thing we do in the function is to test if it both graphs are valid by plugging them into the is_valid function that was already written. There's no point in testing invalid graphs for isomorphism. 

Next, we put each graph through the edge function to obtain dataframes. We then test to see if these two dataframes have the same number of rows or in other words, edges going in one direction. So, although vertex A is going to vertex B with weight w, and vertex B is going to vertex A with weight w, this will show up as A B w and B A w - 2 separate timese in our dataframe. In a special case, it is possible for our dataframe to have 0 rows. This is a result from a graph with a vertex that doesn't link to anything and has no weight. In this case, the only graph it would equal to is a graph with a single vertex of the same name that also doesn't link to anything and has no weight. 

Since our dataframe can't have a negative number of rows, we now consider the remaining case; where our dataframe has 1 or more rows. We use a for loop to run through all the different rows in our first graph's dataframe. We find the entry for the start, end, and weight for the ith row and store them inside variables for ease of use later on. We test to see if the starting vertex in our first graph is a starting vertex anywhere in our second graph. If there is, we store the indexes of those rows in a variable. Now, we check to see if our ending vertex in graph 1 is a ending vertex in any of those rows with the same starting vertex. If there is, there should only be one. We find the weight for that row in our second graph's dataframe with the same starting and ending vertex. Should that weight be equal to the first graph's weight we found earlier, we have an identical edge. Going through all the rows in our first graph's dataframe, if we find an identical edge for all of them in our second graph's dataframe, the two graphs are isomorphic. Remember, we made sure that our two graphs have the same number of edges at the very beginning. As such, there is no way for our second graph to contain all the edges our first graph does and some additional edge (which wouldn't have made the two graphs isomorphic). 

Examples
```{r}
is_isomorphic(graph1, graph1)
```
Let's continue to use graph1 as our example. This will return TRUE since it meets all the vertices, edges, and weights are identical. This is a slightly trivial answer though since graph1 is already viewably identical as itself. 

Let's try two other graphs. 
```{r}
g5 = list(A = list(edges   = c(1L,2L),
                     weights = c(1 ,1 )),
            B = list(edges   = 1L,
                     weights = 1 ))

g6 = list(B = list(edges   = 2L,
                     weights = 1 ),
            A = list(edges   = c(1L,2L),
                     weights = c(1, 1 )))
is_isomorphic(g5,g6)
is_isomorphic(g6,g5)
```
These two should return TRUE as well. Let's use the edge function to make it easier to see why. 
```{r}
edge(g5)
edge(g6)
```
As you can see, g5 has three rows. The first one tells us that A vertex will go to itself with weight 1. In g6, you can see that the third row tells us that this graph has the exact same edge. The same holds for the second and third row when compared to the second and first row respectively. 

Let's add another graph, to compare with. 
```{r}
g7 = list(B = list(edges   = 2L,
                     weights = 1 ),
            A = list(edges   = 1L,
                     weights = 1 ))
is_isomorphic(g7,g5)
is_isomorphic(g5,g7)
is_isomorphic(g6,g7)
is_isomorphic(g7,g6)
```
The first two will return FALSE because while g5 has an edge from vertex A to vertex A with weight 1, graph 7 does not. The same holds for g6 when compared to g7. 

```{r}
g8 = list(A = list(edges   = c(1L,2L),
                     weights = c(1, 2 )),
            B = list(edges   = 1L,
                     weights = 1 ))
g9 = list(A = list(edges   = c(1L,2L),
                         weights = c(2, 1 )),
                B = list(edges   = 1L,
                         weights = 1 ))
is_isomorphic(g8,g9)
```
We get FALSE here too for two reasons. In g8, we have vertex A going to vertex A with weight 1, but in g9, vertex A is going to vertex A with weight 2. The weights differ for one. Another issue is that vertex A is going to vertex B with weight 2 in g8 but weight 1 in g9. 


Connected

The purpose of this funciton is to see if there is a path from a given vertex to another given vertex within a given graph. 

Before running the function, we first run a series of checks on our input variables. We first check if the graph is valid or not. We then check whether or not our edges are null. Then we check if the edges are of the types "logical"--for example, if the user puts in g1, v1, and "TRUE" into the function then the function throws an error that one of the edges specified is not a valid entry. We also check if the edges our numeric variables. If they are, then our function throws an error so that the user can input a character value into v1 and v2. Lastly, if the edges are not in the graph utilized, then the function also throws an error. We implemented a recursion process in running our is connected function. We first call up the function is connected helper, which indexes the vertices that we've already passed when we "walk" from one edge to the other. We use a "seen" array to keep track of the list of vertices the function has already checked. The reasoning for tracking the list of visited vertices is to prevent an infinite recursion when two vertices point to each other. If we have already seen the vertex, then the function passes that vertex and moves on to the next edge. The final output is_connected(g, v1, v2) tells us whether or not edge v1 is connected to edge v2. If v1 and v2 are connected, return TRUE and if v1 and v2 are not connected returns FALSE.  

Continuing to use our graph1 example. 
```{r}
is_connected(graph1, "A", "B")
```
This should give us TRUE since vertex A connects to vertex B with weight 6. 

```{r}
is_connected(graph1, "A", "J")
is_connected(graph1, "J", "A")
```
This should give us FALSE since our function will be making sure that the inputted vertices exist in the inputted graph. So in this case "J" isn't a vertex name in graph1. 

```{r}
is_connected(graph1, 1, 2)
```
This also gives us a FALSE since 1 and 2 are inputted as the vertices, but we want the vertices inputs to be of character value not numeric value. 

```{r}
g10 = list(A = list(edges   = c(1L), 
                      weights = c(1)), 
             B = list(edges   = c(2L), 
                      weights = c(1))) 
is_connected(g10, "A", "B")
```
For this case, we will get FALSE because vertex A only connects to itself and vertex B only connects to itself. They don't connect to each other. 

Shortest Path

The purpose of the shortest_path is to find the shortest path between two vertices inside a given graph through the edges. By shortest path, we mean the sum of weights for the edges in the route taken. 

First, we test to see if the inputted graph is valid. Then if the inputted vertices aren't character types, we stop because it is a bad label for them. It's typically not a very good idea to name something using numerics, since we use numbers when we are subsetting to index. We then check to make sure that the two inputted vertices exist in the inputted graph. 

Now, we apply Dijkstra's Algorithm. Our method begins by picking a starting vertex and mark every other vertex as unvisited. We'll assign each vertex inside our graph two attributes to begin with. The first is distance. To begin with, our distance for each vertex is infinity. The other attribute is previous, which well tell us what the previously visited node is. Starting from our starting vertex, we look at all the vertexes that our starting vertex goes to. We find the one with the shortest distance or weight and pick that one. Now that we are at out second vertex, we mark this vertex as visited. We also change the attribute previous to be the first vertex. Should the distance from the first vertex to our current vertex on the route we took be less than the distance previously assigned to our current vertex, we will replace the old distance with the current one. Note, we never visit an already visited vertex. From our current vertex, we repeat the steps in finding the connected vertex with the shortest distance and replace the previous and distance attribute if appropriate. We repeat these steps until we reach the vertex we want to end at. Since we've recorded our steps on the route, it is easy to go back and map out the shortest_path.

For the corner case where the first vertex goes to itself, Dijkstra's Algorithm won't directly solve the problem. What we do in this case, is look at all the paths from vertices that can go to the first vertex in one step. We use Dijkstra's Algorithm to calculate the shortest route from that. Then, we add the shortest distance to the distance between the penultimate vertex and the last vertex. Also, if possible, we compare the shortest path we found to the distance between the first vertex and itself. 

We implemented the shortest_path function using forloops, while loops and ifelse statements. 

Our graph1 wouldn't illustrate our shortest_path function well. 
We'll us a different one. 
```{r}
g11 = list(A = list(edges   = c(2L,4L,5L),
                    weights = c(1 ,4 ,3 )),
           B = list(edges   = c(1L,4L,5L),
                    weights = c(1 ,4 ,2 )),
           C = list(edges   = c(5L,6L),
                    weights = c(4 ,5 )),
           D = list(edges   = c(1L,2L,5L),
                    weights = c(4 ,4 ,4 )),
           E = list(edges   = c(1L,2L,3L,4L,6L),
                    weights = c(3 ,2 ,4 ,4 ,7 )),
           F = list(edges   = c(3L,5L),
                    weights = c(5 ,7 )))
shortest_path(g11, "B", "F")
```
What we should get back is the path "B", "E", "F". 
