---
output: pdf_document
---
Team4 - Duke Dinosaurs HW 1 Write Up

===================================================================================================

```{r}
source("graph.R")
```
VALID

Our is_valid function will test if the inputted graph object is a list of lists. We also check to see if the names of our vertices are unique, and that each vertex has a list of only edges and weights with edges being integers and weights being numerics. It will also check to make sure that no vertex connects to a nonexistent vertex and that each edge has a weight that is greater than 0. 

We write this function with if statements. Similar to a to-do list, we test conditions to be a valid graph. We start with the broadest conditions, the graph object has to be a list. We go on to test that each vertex is a list that contains edges and weights. The next step is to test if edges and weights are empty. If they aren't empty, the entry must not be NA. Then we test to make sure that edges are integers and weights are numerics. If our edges are greater than the length of our graph, or less than 1, there should be an error. This way, we can ensure that the number of vertices our vertex is connecting to is not more than the number of vertices in our graph and that it's not connecting to nothing. Our weights should not be less than or equal to 0. The number of vertices a vertex connects to should be the same number of weights that vertex holds, so each edge will have a weight. Lastly, we make sure that our vertex is only connecting to another vertex once. 
Let's test out an example graph. 

```{r}
graph1 <- list(A = list(edges   = c(2L, 5L, 6L),
                        weights = c(6, 1, 3)),
               B = list(edges   = c(1L, 3L, 6L),
                        weights = c(6, 4, 13)),
               C = list(edges   = c(2L, 6L),
                        weights = c(4, 7 )),
               D = list(edges   = c(5L, 6L),
                        weights = c(10, 12 )),
               E = list(edges   = c(1L,4L, 6L),
                        weights = c(1, 10, 5)),
               F = list(edges   = c(1L, 2L, 3L, 4L, 5L),
                        weights = c(3, 13, 7, 12, 5)))
```
For simpler viewing purposes: 
> edge(graph1) \newline
   start end weight \newline
     A   B      6 \newline
     A   E      1 \newline
     A   F      3 \newline
     B   A      6 \newline
     B   C      4 \newline
     B   F     13 \newline
     C   B      4 \newline
     C   F      7 \newline
     D   E     10 \newline
     D   F     12 \newline
     E   A      1 \newline
     E   D     10 \newline
     E   F      5 \newline
     F   A      3 \newline
     F   B     13 \newline
     F   C      7 \newline
     F   D     12 \newline
     F   E      5 \newline
 
```{r}
is_valid(graph1)
```

is_valid(graph1) should return TRUE since as you can see from the graphic above, graph1 fits all the criterion of a valid graph. 

> is_valid(graph1) \newline
[1] TRUE \newline

For graphs such as the following, is_valid will return FALSE for various reasons. 

```{r}
bad_g2 <- list(list(edges   = c(1L, 2L),
                     weights = c(1, 1)),
                list(edges   = c(1L, 2L),
                     weights = c(1, 1)))
```
Our is valid function will return this graph input as FALSE because there aren't any unique names for the vertices. In fact, there aren't any names at all. 
> is_valid(bad_g2) \newline
[1] FALSE \newline
Undirected \newline



This function will test to see if a graph is undirected. A graph is undirected if all directed edges within the graph have a directed edge coming back with the same weight. For example, should vertex A connect to B with weight w, B should also connect to A with weight w. 

We first test to see if the inputted graph is valid. 

Afterwards, we utilize a helper function, edge, that will turn a graph into a dataframe, such that each row will represent a directed edge with the first column noting the starting vertex, the second column noting the ending vertex, and the third column noting the weight of that edge. 

We only consider the case where a graph has more than 0 edges. Utilizing a for loop to run through all the rows in the graph's dataframe, we then use if statements to knock out graphs with attriubtes of an undirected graph. We first test to see if there's a row such that the ith row that we are testing will have the starting vertex as the ending vertex and vice-versa. If one does exist, we then test the weight of that row with our ith row. If the weight is the same, then we have a complimentary directed edge for our ith row.  



Isomorphic

The purpose of this function is that given two graph objects, the function will give us TRUE if isomorphic, and FALSE if not. If two graphs are isomorphic, the vertices, edges, and weights are identical. Say we have vertex A in the first graph, and it connects to vertex B with weight w. We should also have a vertex A in the second graph that connects to vertex B with weight w. I call them vertex A and vertex B, because isomorphic graphs will be comparing the names of the vertices, not their position within the graph. This should hold true for the entire graph. 

First thing to note. We have created a helper function called edge. Edge takes a graph object and creates a data frame with three columns, start, end, and weight. The start and end columns will contain the name of the vertices. Weight will be a column of numbers. Each row will represent an edge and list the starting vertex, ending vertex and the weight. 

Aside from that, we utilized if statements and for loops to achieve our purpose. Subsetting and logical operators were also very helpful in eliminating graphs step by step. We worked from the broadest aspects we didn't want and then slowly became more detailed. 

The first thing we do in the function is to test if it both graphs are valid by plugging them into the is_valid function that was already written. There's no point in testing invalid graphs for isomorphism. 

Next, we put each graph through the edge function to obtain dataframes. We then test to see if these two dataframes have the same number of rows or in other words, edges going in one direction. So, although vertex A is going to vertex B with weight w, and vertex B is going to vertex A with weight w, this will show up as A B w and B A w - 2 separate timese in our dataframe. In a special case, it is possible for our dataframe to have 0 rows. This is a result from a graph with a vertex that doesn't link to anything and has no weight. In this case, the only graph it would equal to is a graph with a single vertex of the same name that also doesn't link to anything and has no weight. 

Since our dataframe can't have a negative number of rows, we now consider the remaining case; where our dataframe has 1 or more rows. We use a for loop to run through all the different rows in our first graph's dataframe. We find the entry for the start, end, and weight for the ith row and store them inside variables for ease of use later on. We test to see if the starting vertex in our first graph is a starting vertex anywhere in our second graph. If there is, we store the indexes of those rows in a variable. Now, we check to see if our ending vertex in graph 1 is a ending vertex in any of those rows with the same starting vertex. If there is, there should only be one. We find the weight for that row in our second graph's dataframe with the same starting and ending vertex. Should that weight be equal to the first graph's weight we found earlier, we have an identical edge. Going through all the rows in our first graph's dataframe, if we find an identical edge for all of them in our second graph's dataframe, the two graphs are isomorphic. Remember, we made sure that our two graphs have the same number of edges at the very beginning. As such, there is no way for our second graph to contain all the edges our first graph does and some additional edge (which wouldn't have made the two graphs isomorphic). 




Connected

The purpose of this funciton is to see if there is a path from a given vertex to another given vertex within a given graph. 

Before running the function, we first run a series of checks on our input variables. We first check if the graph is valid or not. We then check whether or not our edges are null. Then we check if the edges are of the types "logical"--for example, if the user puts in g1, v1, and "TRUE" into the function then the function throws an error that one of the edges specified is not a valid entry. We also check if the edges our numeric variables. If they are, then our function throws an error so that the user can input a character value into v1 and v2. Lastly, if the edges are not in the graph utilized, then the function also throws an error. We implemented a recursion process in running our is connected function. We first call up the function is connected helper, which indexes the vertices that we've already passed when we "walk" from one edge to the other. We use a "seen" array to keep track of the list of vertices the function has already checked. The reasoning for tracking the list of visited vertices is to prevent an infinite recursion when two vertices point to each other. If we have already seen the vertex, then the function passes that vertex and moves on to the next edge. The final output is_connected(g, v1, v2) tells us whether or not edge v1 is connected to edge v2. If v1 and v2 are connected, return TRUE and if v1 and v2 are not connected returns FALSE.  


Shortest Path

The purpose of the shortest_path is to find the shortest path between two vertices inside a given graph through the edges. By shortest path, we mean the sum of weights for the edges in the route taken. 

First, we test to see if the inputted graph is valid. Then if the inputted vertices aren't character types, we stop because it is a bad label for them. It's typically not a very good idea to name something using numerics, since we use numbers when we are subsetting to index. We then check to make sure that the two inputted vertices exist in the inputted graph. 

Now, we apply Dijkstra's Algorithm. Our method begins by picking a starting vertex and mark every other vertex as unvisited. We'll assign each vertex inside our graph two attributes to begin with. The first is distance. To begin with, our distance for each vertex is infinity. The other attribute is previous, which well tell us what the previously visited node is. Starting from our starting vertex, we look at all the vertexes that our starting vertex goes to. We find the one with the shortest distance or weight and pick that one. Now that we are at out second vertex, we mark this vertex as visited. We also change the attribute previous to be the first vertex. Should the distance from the first vertex to our current vertex on the route we took be less than the distance previously assigned to our current vertex, we will replace the old distance with the current one. Note, we never visit an already visited vertex. From our current vertex, we repeat the steps in finding the connected vertex with the shortest distance and replace the previous and distance attribute if appropriate. We repeat these steps until we reach the vertex we want to end at. Since we've recorded our steps on the route, it is easy to go back and map out the shortest_path.


